### 2023-03-02 calculus.tex
Resolve/make explicit some todos

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
bidirectional type error localization and recovery. Namely, the calculus provides a judgmental
framework with a unified theory for the design of typing semantics that support ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
\cref{sec:calculus-products}. All rules and theorems may be found in the supplementary appendix,
alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
\cref{sec:calculus-agda}). With products, we then consider how error localization is complicated by
patterned let expressions in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how
the system might be extended to more complex typing features, such as System F-style polymorphism.

@@

% TODO outline the structure of this section at the beginning
% TODO make it clear unmarked/marked languages share same types

@@

  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks}. We call expressions of this language to be
    \emph{marked expressions}, denoted $\ECMV$.
    % TODO more intuition for what marked expressions are?

  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
    inserting error marks where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services. % TODO such as THI.

@@

But what happens when the synthesized type of an expression is \emph{not} consistent with the
expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
\textsc{UASubsume} does not apply, and traditional typing semantics would simply fail the type
checking process at this point. \textsc{MKASubsume} would also not be applicable, leaving marking
undefined in those cases.

% TODO use example more closely related to num here since we haven't talked about booleans yet?

@@

  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynTypeM{\ctx}{x}{\TMV}
  }{MSVar}
\end{mathpar}
% TODO more intuition on why this makes sense?

@@

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynTypeM{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
An unbound variable is marked as such, and since nothing may be said about their types, we
synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
proceed, with usage of the unbound variable permitted in any expression.
% TODO clarify/give intuition as to why this is correct

@@

Unlike numbers and variables, there are explicit synthesis and analysis rules for unmarked lambda
abstractions: % TODO intuition why
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. The analysis rule employs the judgment
$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type}
\cite{cimini:2016}, i.e. it may be considered an arrow type. Defined in
\cref{fig:calculus-matched-arrow}, this notion is purely a technical mechanism to avoid duplication
of rules related to arrow types.

% TODO mention explicitly that by inductive hypothesis, the marking of body cannot fail?
% TODO more intuition on these two judgments
% TODO more intuition on USLam and UALam

@@

  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV}
  }{
    \ctxAnaTypeU{\ctx}{\ECIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV}
  }{UAIf}
\end{mathpar}
%
In synthetic position, conditionals synthesize the \emph{join} of the branch types $\TMV_1$ and
$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. In analytic position, since
there is an expected type for both branches, we shift the blame for any errors to them.

% TODO intuition why we use join

\newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
    \joinsTo{\TNum}{\TNum}{\TNum} \\
    \joinsTo{\TBool}{\TBool}{\TBool} \\
    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}}
  \end{array}\]
  \vspace{-10px}
  \caption{Type join.}
  \label{fig:calculus-type-join}
\end{figure}
% TODO intuition for type join?

@@

% TODO mini conclusion on methodology

@@

\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
  \label{thm:calculus-marking-well-ill-typed}
  \begin{enumerate}
    \item \begin{enumerate}
        \item If $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
          then $\markless{\ECMV}$.
        \item If $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
          then $\markless{\ECMV}$.
      \end{enumerate}

@@

\subsection{Agda Mechanization}
\label{sec:calculus-agda}

@@

The Hazelnut action calculus introduced by \citet{HazelnutPOPL} gives a structure editing scheme
that employs \emph{expression holes}, denoted $\EEHole$, which represent syntactically incomplete
part of the program. Not only are terms always syntactically well-formed, they are always
\emph{well-typed}, ensuring that semantic services are always available. To accomplish this, the
bidirectional action calculus---given by the synthetic action judgment, written
$\ASEAction{\ctx}{\EMV}{\TMV}{\EMV'}{\TMV'}{\AMV}$, and the analytic action judgment, written
$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$---inserts on-demand \emph{error holes}, denoted
$\hhole{\EMV}{}$, which indicate type errors.

% TODO example

@@

% TODO note that we have not verified the metatheory for the typed system

>>>>

The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
bidirectional type error localization with gradual recovery. Namely, the calculus provides a
judgmental framework with a unified theory for the design of typing semantics that support ill-typed
programs. \Cref{sec:calculus-calculus} describes its form and metatheory by example on a small
gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
\cref{sec:calculus-products}. All rules and theorems may be found in the supplementary appendix,
alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
\cref{sec:calculus-agda}). With products, we then consider how error localization is complicated by
patterned let expressions in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how
the system might be extended to more complex typing features, such as System F-style polymorphism.

@@

@@

  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks}. We call expressions of this language to be
    \emph{marked expressions}, denoted $\ECMV$.

  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
    inserting error marks where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services, such as constraint-based inference assistance (as discussed in
\cref{sec:thi}).

@@

But what happens when the synthesized type of an expression is \emph{not} consistent with the
expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
against $\TNum$ in $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
\textsc{UASubsume} does not apply, and traditional typing semantics would simply fail the type
checking process at this point. \textsc{MKASubsume} would also not be applicable, leaving marking
undefined in those cases.

@@

  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynTypeM{\ctx}{x}{\TMV}
  }{MSVar}
\end{mathpar}

@@

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynTypeM{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
An unbound variable is marked as such, and since nothing may be said about their types, we
synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
proceed, with usage of the unbound variable permitted in any expression. \todo{intution for
correctness} \todo{relate to gradual recovery}

@@

Unlike numbers and variables, there are explicit synthesis and analysis rules for unmarked lambda
abstractions. This is because expected input and output types are known, and we may verify that the
type annotation and body match them.
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini:2016}, i.e. it may be
considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
technical mechanism to avoid duplication of rules related to arrow types.

@@

  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV}
  }{
    \ctxAnaTypeU{\ctx}{\ECIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV}
  }{UAIf}
\end{mathpar}
%
In synthetic position, conditionals synthesize the \emph{join} of the branch types $\TMV_1$ and
$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. We choose the ``more
specific'' type of the two\todo{intuition for why this is correct}. In analytic position, since
there is an expected type for both branches, we shift the blame for any errors to them.

\newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
    \joinsTo{\TNum}{\TNum}{\TNum} \\
    \joinsTo{\TBool}{\TBool}{\TBool} \\
    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}}
  \end{array}\]
  \vspace{-10px}
  \caption{Type join.}
  \label{fig:calculus-type-join}
\end{figure}

@@

@@

\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \
  \label{thm:calculus-marking-well-ill-typed}
  \begin{enumerate}
    \item \begin{enumerate}
        \item If $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
          then $\markless{\ECMV}$.
        \item If $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
          then $\markless{\ECMV}$.
      \end{enumerate}

@@

\subsection{Agda Mechanization}
\label{sec:calculus-agda}
% TODO clarify exactly what is/isn't mechanized

@@

The Hazelnut action calculus introduced by \citet{HazelnutPOPL} gives a structure editing scheme
that employs \emph{expression holes}, denoted $\EEHole$, which represent syntactically incomplete
part of the program. Not only are terms always syntactically well-formed, they are always
\emph{well-typed}, ensuring that semantic services are always available. To accomplish this, the
bidirectional action calculus---given by the synthetic action judgment, written
$\ASEAction{\ctx}{\EMV}{\TMV}{\EMV'}{\TMV'}{\AMV}$, and the analytic action judgment, written
$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$---inserts on-demand \emph{error holes}, denoted
$\hhole{\EMV}{}$, which indicate type errors. \todo{example?}

@@
